---
title: "4211pro"
output:
  html_document: default
  pdf_document: default
date: "2024-12-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


```{r}
library(dplyr)
library(tidyverse)
load("/Users/calvinkapral/Downloads/ATP (7).RDA")
library(ggplot2)
library(tidyr)

library(reshape2)
library(haven)
library(pROC)
```


## Problem 1:

Null Hypothesis: There is no association between COVID vaccination uptake and trust in elected officials

Alternative Hypothesis: There is an association between COVID vaccination uptake and trust in elected officials.

Here is the contingency table:
```{r}
data_problem_1 <- ATP %>% mutate(CONF_GOV_grouped = ifelse(CONF_GOV %in% c(1, 2), "Low Trust", "High Trust"))
contingency_table <- table(data_problem_1$COVID_vaccinated, data_problem_1$CONF_GOV_grouped)

contingency_table 

```
Individuals with High Trust in government are more likely to be vaccinated (5759) compared to those with Low Trust (2430).
Among those not vaccinated, High Trust is still more common (1785) compared to Low Trust (442), though the counts are smaller.
Overall, trust in government appears positively associated with COVID vaccination uptake.

Pearson's Chi-square Test
```{r}

chi_sqr <- chisq.test(contingency_table)
chi_sqr
chi_sqr$expected

```
We use chi-square because it is appropriate for testing between categorical variables. Since the p-value is significantly less than 0.05, we reject the null hypothesis. There is a statistically significant association between trust in government officials and COVID vaccination uptake. Individuals with High Trust in government are more likely to be vaccinated compared to those with Low Trust.
\\\newpage

Problem 2
```{r}
filtered_data <- data_problem_1 %>% select(COVID_vaccinated, F_AGECAT, Conf_SCI, trust) %>% drop_na()

age_vacc_rates <- filtered_data %>% group_by(F_AGECAT, COVID_vaccinated) %>% summarise(count = n()) %>% mutate(prop = count / sum(count))

conf_sci_vacc_rates <- filtered_data %>% group_by(Conf_SCI, COVID_vaccinated) %>% summarise(count = n()) %>% mutate(prop = count / sum(count))

ggplot(age_vacc_rates, aes(x = as.factor(F_AGECAT), y = prop, fill = COVID_vaccinated)) + geom_bar(stat = "identity", position = "stack") + labs(title = "Vaccination Rates by Age Group", x = "Age Group", y = "Proportion", fill = "Vaccinated")

ggplot(conf_sci_vacc_rates, aes(x = as.factor(Conf_SCI), y = prop, fill = COVID_vaccinated)) + geom_bar(stat = "identity", position = "stack") + labs(title = "Vaccination Rates by Confidence in Science", x = "Confidence in Science", y = "Proportion", fill = "Vaccinated")

trust_confidence_vacc <- filtered_data %>% group_by(trust, Conf_SCI, COVID_vaccinated) %>% summarise(count = n()) %>% mutate(prop = count / sum(count))

grouped_bar_data <- trust_confidence_vacc %>% filter(COVID_vaccinated == "Y")

ggplot(grouped_bar_data, aes(x = as.factor(Conf_SCI), y = prop, fill = trust)) + geom_bar(stat = "identity", position = "dodge") + labs(title = "Vaccination Rates by Trust and Confidence in Science", x = "Confidence in Science", y = "Proportion Vaccinated", fill = "Trust in Government")


```
Confidence in Science is a critical factor influencing vaccination rates. Higher confidence leads to higher vaccination, irrespective of trust in government.
When confidence in science is low, trust in government plays a more significant role in vaccination decisions.
Age group has minimal impact on vaccination uptake compared to trust and confidence levels.
These findings emphasize the importance of trust in government and confidence in science as key predictors of COVID vaccination uptake.

\\\newpage
Problem 3
```{r}

filtered_data_2 <- data_problem_1 %>% select(COVID_vaccinated, F_ATTEND, F_EDUCCAT, F_PARTY_FINAL, F_RACETHNMOD) %>% drop_na()

education_rates <- filtered_data_2 %>% group_by(F_EDUCCAT, COVID_vaccinated) %>% summarise(count = n()) %>% mutate(prop = count / sum(count))

#1 = college grad, 2 = some college, 3 = hs grad or less

religious_attendance_rates <- filtered_data_2 %>% group_by(F_ATTEND, COVID_vaccinated) %>% summarise(count = n()) %>% mutate(prop = count / sum(count))

#6 = never attends --> #1 = more than once a week
party_vaccination <- filtered_data_2 %>% group_by(F_PARTY_FINAL, COVID_vaccinated) %>% summarise(count = n()) %>% mutate(prop = count / sum(count))

#1 = republican, #2 = democrat, #3 = independent, #4 = other

race_vaccination <- filtered_data_2 %>% group_by(F_RACETHNMOD, COVID_vaccinated) %>% summarise(count = n()) %>% mutate(prop = count / sum(count))

#1 = white, #2 = black, #3 = hispanic, #4 = other, #5 = asian

ggplot(education_rates, aes(x = as.factor(F_EDUCCAT), y = prop, fill = COVID_vaccinated)) + geom_bar(stat = "identity", position = "stack") + labs(title = "Distribution of Education Levels", x = "Education Level", y = "Count")

ggplot(religious_attendance_rates, aes(x = as.factor(F_ATTEND), y = prop, fill = COVID_vaccinated)) + geom_bar(stat = "identity", position = "stack") + labs(title = "Distribution of Religion Attendance", x = "Religion", y = "Count")

ggplot(party_vaccination, aes(x = as.factor(F_PARTY_FINAL), y = prop, fill = COVID_vaccinated)) + geom_bar(stat = "identity", position = "stack") + labs(title = "Distribution of Parties", x = "Party", y = "Count")

ggplot(race_vaccination, aes(x = as.factor(F_RACETHNMOD), y = prop, fill = COVID_vaccinated)) + geom_bar(stat = "identity", position = "stack") + labs(title = "Distribution of Diff Races", x = "Party", y = "Count")

```
As you can see, the GGplots represent the different distributins, with variation among groups.
\\\newpage

Problem 4
Cleaning the Data
```{r}
#First, get rid of 98 and 99 values

data <- ATP #establish data as the call 
data[data == 98 | data == 99] <- NA
```
Get rid of unnessary columns
```{r}
data$QKEY <- NULL
data$INTERVIEW_START <- NULL
data$INTERVIEW_END <- NULL
data$DEVICE_TYPE <- NULL
data$FORM <- NULL
data$WEIGHT_W114 <- NULL
data$WEIGHT_W64_W66_W83_W114 <- NULL
data$WEIGHT_W84_W114 <- NULL

```

\\\newpage
Problem 5
```{r}

data$COVID_vaccinated <- ifelse(data$COVID_vaccinated == "Y", 1, 
                                ifelse(data$COVID_vaccinated == "N", 0, NA))

data <- data %>%
  mutate(across(c(COVID_vaccinated,F_ATTEND
, F_AGECAT, F_GENDER, F_EDUCCAT, F_INC_TIER2, , F_PARTY_FINAL	, F_RACETHNMOD), as.factor))

data$COVID_vaccinated <- relevel(data$COVID_vaccinated, ref = "0") #No as reference for vaccinated
data$F_AGECAT <- relevel(data$F_AGECAT, ref = "4")          # Older as reference
data$F_EDUCCAT <- relevel(data$F_EDUCCAT, ref = "3")        # dint go to college
data$F_PARTY_FINAL	 <- relevel(data$F_PARTY_FINAL	, ref = "1")# "republican" as reference for party
data$F_ATTEND <- relevel(data$F_ATTEND, ref = "1") # reference is never attends
data$F_INC_TIER2 <- relevel(data$F_INC_TIER2, ref = "1")#reference is low
data$F_RACETHNMOD <- relevel(data$F_RACETHNMOD, ref = "2")# reference is white

data_model <- glm(formula = data$COVID_vaccinated ~ data$F_AGECAT + data$F_EDUCCAT + data$F_PARTY_FINAL + data$F_ATTEND+ data$F_INC_TIER2 + data$F_RACETHNMOD,
    family = binomial(link = "logit"), data = data)


summary(data_model)

pchisq(deviance(data_model), df.residual(data_model), lower = FALSE)
exp(confint(data_model))
plot(fitted(data_model), residuals(data_model),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted",
     pch = 20, col = "blue")
abline(h = 0, col = "red", lty = 2)

pearson_residuals <- residuals(data_model, type = "pearson")
overdispersion <- sum(pearson_residuals^2) / df.residual(data_model)
print(overdispersion)


data$predicted <- predict(data_model,newdata = data, type = "response")


# Actual values (binary outcome)
data$actual <- data$COVID_vaccinated


roc_curve <- roc(data$COVID_vaccinated, data$predicted)
auc_value <- auc(roc_curve)
plot(roc_curve, main = "ROC Curve", col = "blue", lwd = 2)

text(0.6, 0.4, paste("AUC =", round(auc_value, 3)), col = "blue")
print(paste("AUC:", round(auc_value, 3)))
```

By calculating for overdispersion, it is less than 1.5 so the binomial model works. (It is 0.99)
There’s a mild curvature in the residuals, especially for larger fitted values, suggesting some non-linearity. However, the deviation is relatively small.
The spread of residuals is fairly consistent but becomes slightly wider and more negative for higher fitted values. This points to a slight skewness or systematic bias.
A few influential points  might exaggerate the skew slightly, but they don't dominate the overall trend.

We found the AUC to be 0.754, which is in the range of moderate discriminative power 
Our model gives us an accurate sense of what is going on between the predictor and response variables
For reference, an AUC of 0.5 would represent a model that is randomly guessing 


\\\newpage
Problem 6

Compared to our reference group, we predict that democrats, asian people, people that never attend religious services, and college graduates are much more likely to be vaccinated. On the contrary, we found that the younger people get, the more unlikely it is that they are vaccinated in comparison to the reference group
We can also see trends within groups. For example, as you attend religious services less and less, your odds of being vaccinated go up. The same thing happens for income and education.






